Countering language attrition with PanLex and the Web of Data

### 2014-01-22 panlex-rdf-paper/2012/MLOD_PanLex/paper_v3.tex
Added the panlex paper to the repo

\documentclass[sw]{iosart2c}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{times}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{dcolumn}
\usepackage{graphicx}
\usepackage{url}
\usepackage{xspace}
\usepackage{color}
\usepackage{eurosym}
\usepackage{todonotes}
\usepackage[breaklinks]{hyperref}
\def\sectionautorefname{Section}
\def\subsectionautorefname{Section}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{array}

%\usepackage{authblk}

\usepackage{textcomp}
\newcolumntype{R}{>{\raggedleft\arraybackslash}X}
\newcolumntype{L}{>{\raggedright\arraybackslash}X}
\newcolumntype{C}{>{\centering\arraybackslash}X}

@@

\usepackage[scaled]{beramono}
\newcommand\Small{\fontsize{9}{9.2}\selectfont}
%\newcommand*\LSTfont{\Small\ttfamily\SetTracking{encoding=*}{-60}\lsstyle}

\usepackage{listings}
%\lstset{language=Python}
%\lstset{%
% morekeywords={Create,View,As,Construct,With,From}
%}%
\lstset{emph={%  
    Create,View,As,Construct,With,From%
    },emphstyle={\color{blue}\bfseries}%
}%

%\usepackage{verbatim}
\lstset{%
    numberbychapter=false,
    numbers=left,
    numberstyle=\tiny,
%    basicstyle=\footnotesize\ttfamily,
    basicstyle=\ttfamily \tiny,
    tabsize=2,
    framexleftmargin=2pt,
    captionpos=b,
    frame=single,
    breaklines=true
}
\lstdefinestyle{rdf}{numberblanklines=true, morekeywords={}}
\lstdefinestyle{sparql}{basicstyle=\ttfamily\scriptsize,numberblanklines=true,
    morekeywords={SELECT,OPTIONAL,FROM,DISTINCT,a,WHERE,FILTER,GROUP,ORDER,LIMIT,BY,IN,AS},
    emph={r,pub,aairObject,verb,person,bday,s,p,o},emphstyle=\textit
}
\lstdefinestyle{turtle}{basicstyle=\ttfamily\tiny ,numberblanklines=true,
    morekeywords={a, @prefix},
    morecomment=[s][\textrm]{<}{>},
    morecomment=[s][\textit]{"}{"},
}

@@

\firstpage{1} \lastpage{5} \volume{1} \pubyear{2012}
\begin{document}
\begin{frontmatter}

%\pretitle{Pretitle}

\title{Countering language attrition with PanLex and the Web of Data}
\runningtitle{Countering language attrition with PanLex and the Web of Data}
%\todo{find a real title}

@@

%patrick.westphal@studserv.uni-leipzig.de
\author[A]{\fnms{Patrick} \snm{Westphal}},
\author[A]{\fnms{Claus} \snm{Stadler}},
\author[B]{\fnms{Jonathan} \snm{Pool}}
%\address[1]{University of Leipzig, patrick.westphal@studserv.uni-leipzig.de}
\address[A]{University of Leipzig, \{pwestphal, cstadler\}@informatik.uni-leipzig.de}
\address[B]{Long Now Foundation, San Francisco, pool@panlex.org}

%\author[1]{Author A\thanks{A.A@university.edu}}
%\author[1]{Author B\thanks{B.B@university.edu}}
%\author[1]{Author C\thanks{C.C@university.edu}}
%\author[2]{Author D\thanks{D.D@university.edu}}
%\author[2]{Author E\thanks{E.E@university.edu}}
%\affil[1]{Department of Computer Science, \LaTeX\ University}
%\affil[2]{Department of Mechanical Engineering, \LaTeX\ University}

%\address[1]{University of Leipzig, Institute of Computer Science, AKSW, patrick.westphal@studserv.uni-leipzig.de}

%\address[1]{University of Leipzig, Institute of Computer Science, AKSW, Postfach 100920, D-04009 Leipzig, Germany}

%\author[A]{Patrick Westphal\thanks{A.A@university.edu}},
%\author[A]{Claus Stadler\thanks{cstadler@informatik.uni-leipzig.de}},
%\author[B]{Jonathan Pool\thanks{todo}}
%\address[A]{University of Leipzig, Institute of Computer Science, AKSW, Postfach 100920, D-04009 Leipzig, Germany\\ E-mail: \{lastname\}@informatik.uni-leipzig.de}
%\address[A]{University of Leipzig, Institute of Computer Science, AKSW, Postfach 100920, D-04009 Leipzig, Germany}
%\address[B]{TODO}

%\author[A]{\fnms{Patrick} \snm{Westphal}}
%\author[B]{\fnms{Claus} \snm{Stadler}}

%\runningauthor{M. Martin et al.}

%\address[B]{University of Leipzig}
%\address[C]{TODO}
%\todo{the e-mail address pattern won't work for both of us}

\begin{abstract}
At present, there are approximately 7,000 living languages in the world.
However, some experts claim that the process of globalization may eventually lead
to the world losing this linguistic diversity.
The vision of the PanLex project is to help save these languages, especially
low-density ones, by allowing them to be intertranslatable and thus to be a part of the Information Age.
For this reason, PanLex gathers and integrates information from thousands of
linguistic resources, such as monolingual dictionaries, bilingual dictionaries, multilingual dictionaries,
glossaries, standards and thesauri.
In this dataset description paper we detail how we transformed this data to
RDF, interlinked it with Lexvo and DBpedia and published it as Linked Data and
via SPARQL.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
% \begin{itemize}
% \item introduce PanLex project
% \item @author: Jonathan
% \item Motivation: About approximately 6900 active languages (\todo{source seems to be http://www.amazon.com/Ethnologue-Languages-M-Paul-Lewis/dp/1556712162}), Existing translation only between relatively few language pairs (dictionaries).
% \item Transitive word-by-word translations 
% * high density and low density languages
% * decline in linguistic diversity
% * These combined forces have led to predictions that something between half and 90%
%   of the world's living languages will die within the next century (Woodbury 2006; UNESCO 2003: 2).
%   Weaker forms of these forces appear to be shrinking the use of medium-density languages in science,
%   diplomacy, business, and other domains (Phillipson 2008).
% * When languages die, they argue, the world loses
%   (1) irreplaceable knowledge of history, medicine, nature, and productive methods encoded in languages'
% * extraction + integration
% \end{itemize}
At present, there are about 7,000 living languages in the world\footnote{See \url{http://www.sil.org/iso639-3/download.asp} for a list of registered languages}.
Nonetheless, some experts claim that processes such as nation-state consolidation and globalization are producing language attrition so rapidly that up to 90\% of all languages alive today will be extinct within a century~\cite{endang_lang}.
Theorists of biolinguistic diversity argue that the loss of
language diversity, the loss of human biological knowledge, and the loss of
species diversity are mutually supportive and thus that language preservation
and revitalization are essential to the preservation of biological diversity~\cite{nettle}.
The vision of the PanLex project is to help save these thousands of
languages, especially those low-density ones that are threatened by extinction,
by supporting their use in global communication. This requires panlingual
translation: translation from any language of the world into any other.
One of the crucial components of panlingual translation of discourses is
panlingual lexical translation. PanLex is designed to support that component.
It documents the known lexical translations (translations of lexemes) among all
languages.

Although a list of all the translations of all words into all other languages
would be large (amounting to trillions of translations), there are more
serious obstacles. The vast majority of these translations are unknown.
Those that are known are not universally agreed on. And lexical polysemy
and ambiguity make the question ``What is the translation of lexeme X into language Y?''
underspecified for practical purposes.
So the PanLex project is designed to permit the application of artificial
intelligence techniques to infer and select translations.

Currently there is a growing community trying to combine linguistic knowledge
with Semantic Web technologies and thereby build a Web of Linguistic Data,
also known as the \emph{Linguistic Linked Open Data (LLOD) cloud}.
The PanLex project is still in its data-acquisition phase, but it has provided a
few web services\footnote{\url{http://panlex.org/try/}} and
APIs\footnote{\url{http://panlex.org/tech/doc/api/}} to access its data.
Since the project shares the idea of open access, we have undertaken the project
of making PanLex a part of the LLOD cloud.

In \autoref{sec:triplificate} we introduce the PanLex dataset, present our
PanLex RDF vocabulary and explain how we transformed the one into the other.
\autoref{sec:linking} is about how we linked to other datasets of the LLOD
cloud, whereas \autoref{sec:publishing} is about the actual dataset publishing.
Usage scenarios are given in \autoref{sec:usage}.
In \autoref{sec:related} we discuss related work, and finally, in
\autoref{sec:conclusion} we conclude our approach and give some hints to future
work.

\section{Triplification of the Raw Data}
\label{sec:triplificate}
In this section we first provide an analysis of the original PanLex dataset.
Subsequently, we introduce our URI and vocabulary design and we explain 
the steps taken to publish the data as RDF.
%The vocabulary 
%afterwards we show how the original data was converted to RDF and published
%via a SPARQL endpoint.
%The interlinking of the created RDF data with other datasets is discussed in the
%following section.

\subsection{Analysis of the Original Dataset}
\label{sec:analysis}
At the core of the PanLex \emph{project}, there is the PanLex
\emph{database} which is created from the imports of thousands of lexical
resources, such as monolingual dictionaries, bilingual dictionaries, multilingual dictionaries, glossaries, standards, and thesauri.
The concrete list of used sources is available online\footnote{\url{http://panlex.org/tech/plrefs.shtml}}.
The data derived from these sources comprises single- and multi-word expressions and meanings assigned to them.
Conceptually, the PanLex database thus ``represents \emph{assertions} about the \emph{meanings} of \emph{expressions}''\footnote{\url{http://panlex.org/tech/doc/design/panlex-db-design.pdf}}.
As of now, the database contains about 20 million meanings and 19 million expressions extracted from about 2,000 sources.
The most important entities and their relations are depicted in \autoref{fig:db-schema} and are explained in more the detail in the following.

%At the core, PanLex is a \emph{database}, served by a PostgreSQL database
%server, that ``represents \emph{assertions} about the \emph{meanings} of
% \emph{expressions}''\footnote{\url{http://panlex.org/tech/doc/design/panlex-db-design.pdf}}.
%As of now, the database contains about 20 million meanings and 19 million
% expressions extracted from about 2,000 sources.

%Assertions are factual claims attributed to a certain source and editor.
%They do not represent absolute truths and as such it is valid that different
% editors make contradictory assertions based on the same sources.

%Expressions are lexemes (dictionary entries) and are related to their
% respective languages.
%Furthermore, expressions have meanings. All expressions sharing the same
% meaning are either translations or synonyms.

%The lexical database of the PanLex project is based on different sources like
% monolingual dictionaries, bilingual dictionaries, multilingual dictionaries, glossaries, standards, and thesauri.
%The list of used sources can also be found on the
% Web\footnote{\url{http://panlex.org/tech/plrefs.shtml}}.

%The actual data derived from these sources comprises single- and multi-word
% expressions and meanings assigned to them.
%The most important entities and their relations are depicted in
% \autoref{fig:db-schema} and are explained in more the detail in the following.

%This denotation process is also reflected in the database design shown below.
%To carry out the goal of a multi-lingual information source a suitable database design is needed.
\begin{figure}
\centering
\includegraphics[width=0.4\textwidth]{images/schema.png}
\caption{The PanLex database schema}
\label{fig:db-schema}
\end{figure}

\begin{itemize}
  %\item A meaning's \emph{approver} is a pair comprised of the editor and the considered information source.
  \item The starting point of the data acquisition is the \emph{approver}
  entity: An editor processes the content of a certain
  mono- or multilingual source as mentioned above and adds it to the PanLex dataset.
  This combination of a user and a source is referred to as an approver.

  \item \emph{Expressions} are lemmas, i.e. dictionary entries. For example,
  \emph{``go''} is a valid expression, whereas \emph{``went''} is not.
  Expressions are always given in a language variety and can only be given once per language variety.
  %They were extracted from the afore-mentioned mono- or bilingual dictionaries
  % or from other sources.
  \item \emph{Languages} in PanLex are identified using ISO 639-3\footnote{\url{http://www.sil.org/iso639-3/codes.asp}} individual and macrolanguage codes, ISO 639-2\footnote{\url{http://www.loc.gov/standards/iso639-2/}} collective codes and ISO 639-5\footnote{\url{http://www.loc.gov/standards/iso639-5/}} codes.
  \item \emph {Language varieties} allow one to make more fine grained
  distinctions within a language. Their codes are composed of the language code
  combined with a PanLex specific identifier.
  For example, ``eng''
  is the ISO 639-3 code for the English language.
  Panlex defines various varieties of it, including ``English'' (eng-000), ``Simple English'' (eng-001) and ``British English'' (eng-005).
  Their labels, when possible, are autonyms, written in the native writing system.
  So, in contrast to mechanisms like IETF BCP 47\footnote{\url{http://tools.ietf.org/html/bcp47}} there is no need for a transcription. %  and one can use variety names like ``{زبان اوستایی}''.
  %Additionally these variety identifiers and labels do not have to be registered like the IETF BCP 47 language subtags\footnote{\url{http://www.iana.org/assignments/language-subtag-registry}}.
  \item \emph{Meanings} in PanLex are entities of which each represents a unique
  possible sense of an expression. 
  %They are allocated as necessary by an editor based on his or her
 %interpretation of expressions.
  Meanings are assigned by editors based on their interpretation of expressions.
  Usually this assignment is done on a per-source basis so that identical meanings
  across multiple sources are not resolved.
  %different
  % meaning identifiers may refer to the same intended sense.
  This means that if there is e.g. a translation of the fruit \emph{``apple''}
  in an English to German dictionary and another translation from English to
  French, these do not necessarily result in a single meaning entity linking to
  all involved languages.
  %the to corresponding expressions in English, German and French.
  Instead, there could be \emph{``apple''} and \emph{``Apfel''} sharing one
  meaning entity and \emph{``apple''} and \emph{``pomme''} sharing another.
%  Editors may add informal definitions to meanings in any language variety.
%  Information sources can be dictionaries, thesauri, etc., and also personal knowledge. 
%  The pair comprised of the editor and the information source is referred to as a meaning's \emph{approver}.
  \item \emph{Definitions} are optional descriptions of a meaning.
  They are given in a certain language variety.
  A description of the verb \emph{browse} for example could be \emph{``move or surf through various files on a computer, the Internet, etc.''}, marked as a definition in the ``English'' language variety.
  \item \emph{Denotations} are entities that relate expressions to meanings and
  may optionally carry annotations in form of sets of key value pairs. For
  instance, an English expression \emph{pig}, when referring to
  \emph{police officer}, could be annotated with \emph{pragmatics=vulgar}. 
  Furthermore, denotations can be tagged with part-of-speech tags, such as word
  classes, selected from a closed list based on the Open Lexicon Interchange Format (OLIF) standard\footnote{\url{http://www.olif.net/}}.
  For example, \emph{fall} can be a verb or a noun for autumn. 
  Homonyms are those expressions that are connected to multiple meanings.
  \item \emph{Users} have editorial privileges over the language varieties and the approvers that they define.
  \item \emph{Licenses} are also considered by the PanLex project.
  At present there are ten different license categories an approver can be annotated with.
  They are \emph{public domain}, \emph{Creative Commons}, \emph{request} (meaning that one has to ask the author of the resource), \emph{GNU General Public License}, \emph{GNU Lesser General Public License}, \emph{GNU Free Documentation License}, \emph{MIT License}, \emph{copyright} (stating that there is a certain copyright holder), \emph{other} and \emph{unknown}.
\end{itemize}

An overview of the number of instances per entity in the current PanLex database is given in~\autoref{fig:plx-entity-counts}.
\begin{table}
\centering
\begin{scriptsize}
\begin{tabular}{lr}
\toprule
Entity & Instances \\
%\midrule
Approvers & 3,905 \\ % SELECT COUNT(ex) FROM ex;
Expressions & 18,580,594 \\ % SELECT COUNT(ex) FROM ex;
%\midrule
Languages   & 7,839 \\ % SELECT COUNT(lc) FROM lc; -- all language codes (ISO 639-3, ISO 639-2, ISO 639-5) known in the PanLex db
% FIXME: there should not be more languages than language varieties
%\midrule
Language Varieties & 7,248 \\ % SELECT COUNT(lv) FROM lv;
%\midrule
Meanings    & 20,023,427 \\ % SELECT COUNT(mn) FROM mn;
%\midrule
Definitions & 2,522,605 \\ % SELECT COUNT(df) FROM df;
%\midrule
Denotations & 50,803,243 \\ % SELECT COUNT(dn) FROM dn;
%\midrule
Users       & 7 \\ % SELECT COUNT(us) FROM us;
%\midrule
Licenses    & 10 \\ % SELECT COUNT(li) FROM apli;
\bottomrule
\end{tabular}
\end{scriptsize}
\caption{Number of entities in the PanLex database}
\label{fig:plx-entity-counts}
\end{table}

%\todo{Possibly also: approver editors, language variety editors, exemplar/approved characters, approver varieties}

%The same expression may be mentioned in multiple information sources.
%However, an expression's possible meanings depend on the user's interpretion of them in context of such a source.
%For this reason, every meaning depends on an approver.
%\emph{Denotations} capture the mappings between expressions and their possible meanings.  
Note that \emph{approver} combines a user with an \emph{information source},
however the \emph{information source} is not modeled as a distinct entity.
Also, the information of whether or not two meanings with different approvers
are the same is not being captured.

%These sources are fundamental for the PanLex database design and are aggregated in the approver table.
%According to these approvers there are special meanings that are represented by expressions.
%%Meanings are always related to an approver, which captures how a user interprets an expression information source.

%So, if there are two expressions, they can either be synonyms (if they are given in the same language) or translations of each other.
%\\
%Expressions are lexemes that can be seen as entries of a lexicon or dictionary of a given language.
%They are represented in their lemma or dictionary form.

%These expressions can be given in different language varieties allowing a meaning preserving translation between these varieties.
%An expression is given in a language variety.
%Such a variety can be a standard variety, a dialect, a script-based variety a controlled technical variety etc.
%Examples would be English, Simple English, British English and American English.\\

%  Ghidorah, the Three-Headed Monster | Ghidorah
%  describe or draw attention to (a product, service, or event) in a public medium in order to promote sales or attendance | advertise
%  move or surf through various files on a computer, the Internet, etc. | browse

>>>>

\documentclass[sw]{iosart2c}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{times}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{dcolumn}
\usepackage{graphicx}
\usepackage{url}
\usepackage{xspace}
\usepackage{color}
\usepackage{eurosym}
\usepackage{todonotes}
\usepackage[breaklinks]{hyperref}

\def\sectionautorefname{Section}
\def\subsectionautorefname{Section}

\usepackage{multirow}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{array}
\usepackage{textcomp}

\newcolumntype{R}{>{\raggedleft\arraybackslash}X}
\newcolumntype{L}{>{\raggedright\arraybackslash}X}
\newcolumntype{C}{>{\centering\arraybackslash}X}

@@

\usepackage[scaled]{beramono}
\newcommand\Small{\fontsize{9}{9.2}\selectfont}

\usepackage{listings}
%\lstset{language=Python}
%\lstset{%
% morekeywords={Create,View,As,Construct,With,From}
%}%
\lstset{emph={%  
    Create,View,As,Construct,With,From%
    },emphstyle={\color{blue}\bfseries}%
}%
\lstset{%
    numberbychapter=false,
    numbers=left,
    numberstyle=\tiny,
%    basicstyle=\footnotesize\ttfamily,
    basicstyle=\ttfamily \tiny,
    tabsize=2,
    framexleftmargin=2pt,
    captionpos=b,
    frame=single,
    breaklines=true
}
\lstdefinestyle{rdf}{numberblanklines=true, morekeywords={}}
\lstdefinestyle{sparql}{basicstyle=\ttfamily\scriptsize,numberblanklines=true,
    morekeywords={SELECT,OPTIONAL,FROM,DISTINCT,a,WHERE,FILTER,GROUP,ORDER,LIMIT,BY,IN,AS},
    emph={r,pub,aairObject,verb,person,bday,s,p,o},emphstyle=\textit
}
\lstdefinestyle{turtle}{basicstyle=\ttfamily\tiny ,numberblanklines=true,
    morekeywords={a, @prefix},
    morecomment=[s][\textrm]{<}{>},
    morecomment=[s][\textit]{"}{"},
}

@@

\firstpage{1} \lastpage{5} \volume{1} \pubyear{2012}
\begin{document}
\begin{frontmatter} 
\title{Countering language attrition with PanLex and the Web of Data}
\runningtitle{Countering language attrition with PanLex and the Web of Data}

@@

\author[A]{\fnms{Patrick} \snm{Westphal}},
\author[A]{\fnms{Claus} \snm{Stadler}},
\author[B]{\fnms{Jonathan} \snm{Pool}}
\address[A]{University of Leipzig, \{pwestphal, cstadler\}@informatik.uni-leipzig.de}
\address[B]{Long Now Foundation, San Francisco, pool@panlex.org}

\begin{abstract}
At present, there are approximately 7,000 living languages in the world.
However, some experts claim that the process of globalization may eventually lead to the world losing this linguistic diversity.
The vision of the PanLex project is to help save these languages, especially low-density ones, by allowing them to be intertranslatable and thus to be a part of the Information Age.
Semantic Web technologies can support achieving this goal, for reasons such as
their capabilities of flexibly representing, interlinking and reasoning with 
data, in our case particularly linguistic resources and annotations.
%Using Semantic Web technologies can support achieving this goal, especially due
% to inference capabilities and the interlinking of the PanLex data with other data sources.
Conversely, an RDF version of PanLex makes a significant contribution towards
improving the coverage of the Linguistic Web of Data, as to the best of our
knowledge there exists no large scale Linked Data data set for panlingual translation of
non-mainstream languages.
In this dataset description paper we detail how we transformed the data of the PanLex project to RDF,
established conformance with the lemon and GOLD data models,
interlinked it with Lexvo and DBpedia, and published it as Linked Data and via
SPARQL.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
At present, there are about 7,000 living languages in the world\footnote{See \url{http://www.sil.org/iso639-3/download.asp} for a list of registered languages}.
Nonetheless, some experts claim that processes such as nation-state consolidation and globalization are producing language attrition so rapidly that up to 90\% of all languages alive today will be extinct within a century~\cite{endang_lang}.
Theorists of biolinguistic diversity argue that the loss of language diversity, the loss of human biological knowledge, and the loss of species diversity are mutually supportive and thus that language preservation and revitalization are essential to the preservation of biological diversity~\cite{nettle}.
The vision of the PanLex project is to help save these thousands of languages, especially those low-density ones that are threatened by extinction, by supporting their use in global communication.
This requires panlingual translation: translation from any language of the world into any other.
One of the crucial components of panlingual translation of discourses is panlingual lexical translation.
PanLex is designed to support that component by making use of several thousand
sources and artificial intelligence approaches.
%It documents the known lexical translations (translations of lexemes) among all
%languages.

Apart from using PanLex's data for inferring lexical translations, a
further direction of work is to generate added value by enriching this data with
knowledge from other (linguistic) sources.
Grounding on the idea of a Semantic Web, a complete stack of technologies has
been devised and standardized by the
\emph{World Wide Web Consortium}\footnote{\url{http://www.w3.org/standards/semanticweb/}}
supporting the definition of a machine readable and interpretable Linked Data
network, which has led to emergence of the \emph{Web of Data}.
%Technologies that support this aim are already available and in use in the so
%called \emph{Web of Data}.
Currently there is a growing community working on leveraging these Semantic Web
technologies for linguistic knowledge and thereby building a Web of Linguistic
Data, also known as the \emph{Linguistic Linked Open Data (LLOD) cloud}.

Our steps to connect PanLex to this Linked Data network are as follows.
In \autoref{sec:triplificate} we introduce the PanLex dataset, present our
PanLex RDF vocabulary, explain how we transformed the one into the other, and
how we established conformance with additional data models.
\autoref{sec:linking} is about how we linked to other datasets of the LLOD cloud,
whereas \autoref{sec:publishing} is about the actual dataset publishing.
Usage scenarios are given in \autoref{sec:usage}.
In \autoref{sec:related} we discuss related work, and finally, in \autoref{sec:conclusion} we conclude our approach and give some hints to future work.

\section{Triplification of the Raw Data}
\label{sec:triplificate}
In this section, we first provide an analysis of the PanLex dataset.
Subsequently, we introduce our URI and vocabulary design which closely
resembles PanLex's original conceptual model.
%, which was developed independently from other ontology engineering approaches.
Afterwards, we briefly describe how we classified PanLex's
instance data using additional data models. Finally, we explain the steps
taken to transform the data to RDF.

%As PanLex defines its own  conceptual model

\subsection{Analysis of the Original Dataset}
\label{sec:analysis}
At the core of the PanLex \emph{project}, there is the PanLex \emph{database} which is created from the imports of thousands of lexical resources, such as mono- and multilingual dictionaries, glossaries, standards, and thesauri.
The concrete list of used sources is available online\footnote{\url{http://panlex.org/tech/plrefs.shtml}}.
The data derived from these sources comprise single- and multi-word expressions and meanings assigned to them.
Conceptually, the PanLex database thus ``represents \emph{assertions} about the \emph{meanings} of \emph{expressions}''\footnote{\url{http://panlex.org/tech/doc/design/panlex-db-design.pdf}}.
As of now, the database contains about 20 million meanings and 19 million expressions extracted from about 2,000 sources.
The most important entities and relations of PanLex's conceptual model are
depicted in \autoref{fig:db-schema} and are explained in more the detail in the following.

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{images/schema.png}
  \caption{The PanLex database schema}
  \label{fig:db-schema}
\end{figure}

\begin{itemize}
  \item The starting point of the data acquisition is the \emph{approver} entity: An editor processes the content of a certain mono- or multilingual source as mentioned above and adds it to the PanLex dataset.
    This combination of a user and a source is referred to as an approver.
  \item \emph{Expressions} are lemmas, i.e. dictionary entries.
    For example, \emph{``go''} is a valid expression, whereas \emph{``went''} is not.
    Expressions are always given in a language variety and can only be given once per language variety.
  \item \emph{Languages} in PanLex are identified using ISO 639-3\footnote{\url{http://www.sil.org/iso639-3/codes.asp}} individual and macrolanguage codes, ISO 639-2\footnote{\url{http://www.loc.gov/standards/iso639-2/}} collective codes and ISO 639-5\footnote{\url{http://www.loc.gov/standards/iso639-5/}} codes.
  \item \emph {Language varieties} allow one to make more fine grained distinctions within a language.
    Their codes are composed of the language code combined with a PanLex specific identifier.
    For example, ``eng'' is the ISO 639-3 code for the English language.
    Panlex defines various varieties of it, including ``English'' (eng-000), ``Simple English'' (eng-001) and ``British English'' (eng-005).
    Their labels, when possible, are autonyms, written in the native writing system.
    So, in contrast to mechanisms like IETF BCP 47\footnote{\url{http://tools.ietf.org/html/bcp47}} there is no need for a transcription.
  \item \emph{Meanings} in PanLex are entities of which each represents a unique possible sense of an expression.
    Meanings are assigned by editors based on their interpretation of expressions.
    Usually this assignment is done on a per-source basis so that identical meanings across multiple sources are not resolved.
    This means that if there is e.g. a translation of the fruit \emph{``apple''} in an English to German dictionary and another translation from English to French, these do not necessarily result in a single meaning entity linking to all involved languages.
    Instead, there could be \emph{``apple''} and \emph{``Apfel''} sharing one meaning entity and \emph{``apple''} and \emph{``pomme''} sharing another.
  \item \emph{Definitions} are optional descriptions of a meaning.
    They are given in a certain language variety.
    A description of the verb \emph{browse} for example could be \emph{``move or surf through various files on a computer, the Internet, etc.''}, marked as a definition in the ``English'' language variety.
  \item \emph{Denotations} are entities that relate expressions to meanings and may optionally carry annotations in form of sets of key value pairs.
    For instance, an English expression \emph{pig}, when referring to \emph{police officer}, could be annotated with \emph{pragmatics=vulgar}.
    Furthermore, denotations can be tagged with part-of-speech tags, such as word classes, selected from a closed list based on the Open Lexicon Interchange Format (OLIF) standard\footnote{\url{http://www.olif.net/}}.
    %For example, \emph{fall} can be a verb or a noun for autumn.
    %Homonyms are those expressions that are connected to multiple meanings.
  \item \emph{Users} have editorial privileges over the language varieties and the approvers that they define.
  \item \emph{Licenses} are also considered by the PanLex project.
    At present there are ten different license categories an approver can be annotated with.
    They are \emph{public domain}, \emph{Creative Commons (CC)}, \emph{request} (meaning that one has to ask the author of the resource), \emph{GNU General Public License (GPL)}, \emph{GNU Lesser General Public License (LGPL)}, \emph{GNU Free Documentation License (FDL)}, \emph{MIT License}, \emph{copyright} (stating that there is a certain copyright holder), \emph{other} and \emph{unknown}.
    The license distribution is shown in \autoref{tbl:plx-license-counts}.
\end{itemize}

\begin{table}
\centering
\begin{scriptsize}
\begin{tabular}{lrclrclr}
License          & Count &&
License          & Count &&
License          & Count \\
\cline{1-2} \cline{4-5} \cline{7-8}
\emph{unknown}   &  1886 && % 1st
\emph{PD}        & 123 && % 2nd
\emph{LGPL}      &     9 \\ % 3rd
\emph{copyright} &  1212 && % 1st
\emph{other}     &   104 && % 2nd
\emph{request}   &     5 \\ % 3rd
\emph{CC}        &   335 && % 1st
\emph{MIT}       &    32 && % 2nd
\\
\emph{GPL}       &   172 && % 1st
\emph{FDL}       &    24 && % 2nd
\end{tabular}
\end{scriptsize}
\caption{Number of approvers using a certain license}
\label{tbl:plx-license-counts}
\end{table}

An overview of the number of instances per entity in the current PanLex database is given in~\autoref{fig:plx-entity-counts}.
\begin{table}
  \centering\begin{scriptsize}
  \begin{tabular}{lrclr}
    Entity             & Instances  &&
    Entity             & Instances  \\
    \cline{1-2} \cline{4-5}
    Denotations        & 50,803,243 &&  % SELECT COUNT(dn) FROM dn;
    Language Varieties &      7,248 \\  % SELECT COUNT(lv) FROM lv;
    Meanings           & 20,023,427 &&  % SELECT COUNT(mn) FROM mn;
    Approvers          &      3,905 \\  % SELECT COUNT(ex) FROM ex;
    Expressions        & 18,580,594 && % SELECT COUNT(ex) FROM ex;
    Licenses           &         10 \\ % SELECT COUNT(li) FROM apli;
    Definitions        &  2,522,605 && % SELECT COUNT(df) FROM df;
    Users              &          7 \\ % SELECT COUNT(us) FROM us;
    Languages          &      7,839 && % SELECT COUNT(lc) FROM lc; -- all language codes (ISO 639-3, ISO 639-2, ISO 639-5) known in the PanLex db
  \end{tabular}
  \end{scriptsize}
  \caption{Number of entities in the PanLex database}
  \label{fig:plx-entity-counts}
\end{table}

Note that \emph{approver} combines a user with an \emph{information source}, however the \emph{information source} is not modeled as a distinct entity.
Also, the information of whether or not two meanings with different approvers are the same is not being captured.

\subsection{The PanLex vocabulary}
\label{sec:vocabulary}
The entities and relations of the schema described in the previous section serve as the base for the development of the PanLex RDF vocabulary.
In general, all PanLex RDF resources reside in the namespace \texttt{\small <http://ld.panlex.org/plx/>}, abbreviated with \texttt{\small plx}.
An example of the resulting ontology is depicted in \autoref{fig:vocabulary} and summarized as follows:
Unless otherwise noted, the URIs of instances of PanLex classes follow the pattern \emph{plx:\{className\}/\{id\}}, where \{className\} is spelled in lower camel case and the \{id\} is the primary key of the corresponding database table.

\begin{itemize}
  \item Expressions are modeled as instances of the class \texttt{\small plx:Expression}.
    Their original and normalized textual representations become the values of the properties \texttt{\small rdfs:label} and \texttt{\small plx:degradedText}, respectively.
    Their corresponding language variety is stated using \texttt{\small plx:languageVariety}.
  \item For language and language varieties the classes \texttt{\small plx:Language} and \texttt{\small plx:LanguageVariety} are introduced.
    \emph{ISO 639-1} and \emph{ISO 639-3 codes} become instances of the classes \texttt{\small plx:Iso639-1Code} and \texttt{\small plx:Iso639-3Code}.
  \item The RDF analog of the PanLex \emph{meaning} is the \texttt{\small plx:Meaning}.
    Entities of this class may have an identifier assigned with the \texttt{\small plx:identifier} property pointing to an \texttt{\small xsd:string} literal.
    Meanings may also have \emph{definitions}, entities of the \texttt{\small plx:Definition} class, giving a textual representation (\texttt{\small rdfs:label}) in a certain language variety (\texttt{\small plx:languageVariety}).
  \item Following the semantics of the PanLex database, meanings and expressions are linked via \emph{denotations}.
    These are entities of the \texttt{\small plx:Denotation} class pointing to meanings and expressions via the properties \texttt{\small plx:denotationMeaning} and \\ \texttt{\small plx:denotationExpression}.
    Denotations may also have a word class assigned to them.
    This can be achieved with the denotation's \texttt{\small plx:wordClass} property pointing to a \texttt{\small plx:WordClass} entity.
  \item All approvers share the \texttt{\small plx:Approver} class.
    The characteristics of an approver are described using mainly triples with literal objects.
    These are for example \texttt{\small dc:title} to assign the title of a source, \texttt{\small dc:creator} to give an \texttt{\small xsd:string} containing the author's name.
\end{itemize}
Since the PanLex project compiled its database by extracting data from different sources, the licenses of these sources were also considered.
At present, we support the different license categories given in the database by creating resources of the \texttt{\small plx:License} class.
